---
title: "Project 3"
format: md
editor: visual
---

## 1. Introduction and data

The goal of this project is to explore patterns in tech salaries based on job titles, experience level, and company characteristics. The data was collected from aijobs.net, where individuals voluntarily submit their salary and job information. The dataset is public, downloadable as a clean CSV file, and includes fields such as job title, experience level, employment type, company size, location, and salary figures. This analysis focuses on identifying correlations between salaries and job-related factors, with the aim to potentially build a model that can predict salary ranges based on observable characteristics. The main variables of interest are listed below.

-   **experience_level**: The experience level in the job during the year.
    -   EN - Entry level
    -   MI – Mid-level
    -   SE – Senior-level
    -   EX – Executive-level
-   **employment_type**: The type of employment:
    -   PT – Part-time
    -   FT – Full-time
    -   CT – Contract
    -   FL – Freelance
-   **job_title**: The role worked during the year.
-   **salary_in_usd**: The salary in USD.
-   **remote_ratio**: The overall amount of work done remotely:
    -   0 – No remote work (less than 20%)
    -   50 – Partially remote/hybrid
    -   100 – Fully remote (more than 80%)
-   **company_location**: The country of the employer’s main office or contracting branch.
-   **company_size**: The average number of people that worked for the company during the year:
    -   S – Small (\< 50 employees)
    -   M – Medium (50–250)
    -   L – Large (\> 250)

### Exploratory Data Analysis (EDA)

#### Top 20 Most Frequent Job Titles by Experience Level

The plot below shows the top 20 most common job titles in the dataset, broken down by experience level.

```{r}
#Splitting dataset

library(tidyverse)

salaries = read.csv("../Project1/data/salaries.csv")

set.seed(1)

#70% training, 15% validating and testing set
train_split = sample(nrow(salaries), size = 0.7 * nrow(salaries))
train_set = salaries[train_split, ]

#splits rest of the 30%
remaining = salaries[-train_split, ] 
valid_split <- sample(nrow(remaining), size = 0.5 * nrow(remaining))

valid_set = remaining[valid_split, ]
test_set = remaining[-valid_split, ]

library(ggthemes)

#top 20 most frequent positions by count (list of job titles)
top_20 = train_set %>%
  count(job_title, sort = TRUE) %>%
  slice_max(n, n = 20) %>%
  pull(job_title) 

#filter training data for only top 20 jobs(dataset)
top_20_set =  train_set %>% 
  filter(job_title %in% top_20)

#horizontal barplot

ggplot(top_20_set, aes(x = fct_rev(fct_infreq(job_title)), fill = experience_level)) + 
  geom_bar() + 
  coord_flip() + 
  labs(
    x = "Job Title",
    y = "Count",
    title = "Top 20 Most Frequent Job Titles by Experience Level",
    fill = "Experience Level"
  ) + 
  theme_clean() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
```

#### Highest Average Salaries Among Frequent Job Titles

This chart shows the average salary for the most frequently reported job titles.

```{r}
#average salary for each of the top 20
avg_salary_top_20 = top_20_set %>%
  group_by(job_title) %>%
  summarise(avg_salary = mean(salary_in_usd, na.rm = TRUE)) %>%
  arrange(desc(avg_salary)) %>%
  slice_max(avg_salary, n = 20) #show only the top 10

ggplot(avg_salary_top_20, aes(x = fct_reorder(job_title, -avg_salary), y = avg_salary, fill = job_title)) + 
  geom_col() + 
  labs(
    title = "Highest Average Salaries for Frequent Job Titles",
    x = "Job Title",
    y = "Average Salary (USD)"
  ) +
  theme_clean() + theme(axis.text.x = element_text(angle = 45, hjust = 1),legend.position = "none")
```

#### Global Distribution of Tech Salaries

The map below provides a global view of salary submissions by employee residence.

```{r}
# Calculate average salary by employee residence
# Calculate average salary by employee residence
avg_sal_country <- train_set %>%
  group_by(employee_residence) %>%
  filter(n() >= 5) %>%
  summarise(avg_salary = mean(salary_in_usd, na.rm = TRUE), .groups = "drop") %>%
  mutate(country_name = countrycode(employee_residence, "iso2c", "country.name"))    %>% arrange(desc(avg_salary))

# Fix names for countries not matched by countrycode
avg_sal_country$country_name[avg_sal_country$employee_residence == "XK"] <- "Kosovo"
avg_sal_country$country_name[avg_sal_country$employee_residence == "US"] <- "USA"
avg_sal_country$country_name[avg_sal_country$employee_residence == "GB"] <- "UK"
avg_sal_country$country_name[avg_sal_country$employee_residence == "CZ"] <- "Czech Republic"

# Load world map data
world_map = map_data("world")

# Merge map with salary data
merged = left_join(world_map, avg_sal_country, by = c('region' = 'country_name'))

# Plot map
ggplot(data = merged, aes(x = long, y = lat, group = group, fill = avg_salary)) +
  geom_polygon(color = "white") +
  scale_fill_viridis_c(
    option = "C",
    na.value = "gray90",
    breaks = seq(50000, 300000, by = 50000),
    labels = label_comma()
  ) +
  labs(
    title = "Global Distribution of Average Tech Salaries by Employee Residence",
    fill = "Avg Salary (USD)"
  ) +
  theme_map() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```

#### Experience Level by Company Size

This heatmap visualizes the frequency of salary reports across different experience levels and company sizes.

```{r}
#creates table
heat_data = train_set %>%
  count(company_size, experience_level)

table(train_set$company_size)

ggplot(heat_data, aes(x = company_size, y = experience_level, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), color = "black", size = 4) +
  labs(title = "Experience Level by Company Size",
       x = "Company Size",
       y = "Experience Level") +
  scale_fill_gradient(name='Count', low = "lightblue", high = "darkblue")+theme_pander()
```

## 2. Methodology/Model

## 📊 Method and Model

For this analysis, I used Light Gradient Boosting Machine (LightGBM), a gradient boosting algorithm, to predict salaries based on categorical variables such as job title, experience level, and company size. Gradient boosting models like LightGBM are effective in capturing non-linear relationships and interactions between features. LightGBM builds an ensemble of decision trees to make predictions, growing trees in a leaf-wise manner to maximize loss reduction at each step. It is designed for high efficiency and accuracy, particularly with large datasets and numeric features.

To preprocess the categorical variables, I applied target encoding. This technique replaces each category in a categorical feature with the mean of the target variable (salary) for that category. For example, if the `experience_level` feature includes categories like `"Junior"` and `"Executive"`, target encoding would replace `"Junior"` with the average salary of all junior-level employees, and `"Executive"` with the average salary of executives. This method is helpful because it transforms high-cardinality categorical variables into numerical representations that gradient boosting models can interpret effectively.

The data was split into 70% training, 15% validation, and 15% test sets to ensure robust model evaluation. The model was trained with a regression objective and tuned using different combinations of hyperparameters to improve performance.

Model performance was evaluated using Root Mean Squared Error (RMSE), which measures the average magnitude of prediction error in the same units as the target variable (USD). For instance, an RMSE of 10,000 indicates that on average, the model's salary predictions were off by approximately \$10,000.

```{r}
#convert categorical variables as factors
train_set$experience_level <- as.factor(train_set$experience_level)
train_set$company_size <- as.factor(train_set$company_size)
train_set$employment_type <- as.factor(train_set$employment_type)
train_set$employee_residence <- as.factor(train_set$employee_residence)
train_set$company_location <- as.factor(train_set$company_location)
train_set$job_title <- as.factor(train_set$job_title)
train_set$remote_ratio <- as.factor(train_set$remote_ratio)

```

```{r}
library(dplyr)
# Target Encoding function
target_encode <- function(data, cat_col, target_col) {
  # Calculate the mean salary for each category
  encoded_values <- data %>%
    group_by_at(cat_col) %>%
    summarise(encoding = mean(!!sym(target_col), na.rm = TRUE)) %>% #for each category, calculates mean of salary
    ungroup()                                                       #!!sym() convert column name into a variable
  
  # Merge encoded values back into the original dataset
  data <- data %>%
    left_join(encoded_values, by = setNames(cat_col, cat_col)) %>%
    mutate(!!sym(cat_col) := encoding) %>%
    select(-encoding) # Remove the encoding column
  
  return(data)
}
```

### Training

```{r}
# Perform target encoding on categorical variables
train_set_encoded <- train_set %>%
  target_encode('experience_level', 'salary_in_usd') %>%
  target_encode('company_size', 'salary_in_usd') %>%
  target_encode('employment_type', 'salary_in_usd') %>%
  target_encode('employee_residence', 'salary_in_usd') %>%
  target_encode('company_location', 'salary_in_usd') %>%
  target_encode('job_title', 'salary_in_usd') %>%
  target_encode('remote_ratio', 'salary_in_usd')

# Set up training data for LightGBM
train_data <- as.matrix(train_set_encoded %>% select(-salary_in_usd))
#target column
train_label <- train_set_encoded$salary_in_usd

library(lightgbm)
# Convert to LightGBM dataset format
dtrain <- lgb.Dataset(data = train_data, label = train_label)

# parameters for LightGBM
params <- list(
  objective = "regression",
  metric = "rmse",
  num_leaves = 31, #controls complexity of trees(larger means more complex model)
  learning_rate = 0.05, #step size at each iteration
  feature_fraction = 0.9, #fraction of features to use during training to prevent overfitting
  bagging_fraction = 0.8, #fraction of data to use during each training iteration (random sampling)
  bagging_freq = 5,
  num_iterations = 1000 #number of boosting iterations (trees)
)

# Train the LightGBM model
model <- lgb.train(params, dtrain, 1000)

# predictions
preds <- predict(model, train_data)

# Calculate R-squared
mse <- mean((train_label - preds)^2)

rmse <- sqrt(mse)

print(paste("Training MSE: ", mse))

print(paste("Training RMSE: ", rmse))
```

```{r}
# Predicted vs Actual plot
plot(train_label, preds, 
     xlab = "Actual Salary (USD)", 
     ylab = "Predicted Salary (USD)", 
     main = "Predicted vs Actual Salary",
     pch = 16, cex = 0.5, col = rgb(0,0,1,0.5))
abline(0, 1, col = "red", lwd = 2)

```

### Validation

```{r}
#convert categorical variables as factors
valid_set$experience_level <- as.factor(valid_set$experience_level)
valid_set$company_size <- as.factor(valid_set$company_size)
valid_set$employment_type <- as.factor(valid_set$employment_type)
valid_set$employee_residence <- as.factor(valid_set$employee_residence)
valid_set$company_location <- as.factor(valid_set$company_location)
valid_set$job_title <- as.factor(valid_set$job_title)
valid_set$remote_ratio <- as.factor(valid_set$remote_ratio)
```

```{r}
valid_set_encoded <- valid_set %>%
  target_encode('experience_level', 'salary_in_usd') %>%
  target_encode('company_size', 'salary_in_usd') %>%
  target_encode('employment_type', 'salary_in_usd') %>%
  target_encode('employee_residence', 'salary_in_usd') %>%
  target_encode('company_location', 'salary_in_usd') %>%
  target_encode('job_title', 'salary_in_usd') %>%
  target_encode('remote_ratio', 'salary_in_usd')

#matrix for validation data
valid_data <- as.matrix(valid_set_encoded %>% select(-salary_in_usd))

valid_label <- valid_set_encoded$salary_in_usd

valid_preds <- predict(model, valid_data)
```

```{r}
# Calculate R-squared
mse2 <- mean((valid_label - valid_preds)^2)

rmse2 <- sqrt(mse2)

print(paste("Validation MSE: ", mse2))

print(paste("Validation RMSE: ", rmse2))
```

### Testing

```{r}
test_set_encoded <- test_set %>%
  target_encode('experience_level', 'salary_in_usd') %>%
  target_encode('company_size', 'salary_in_usd') %>%
  target_encode('employment_type', 'salary_in_usd') %>%
  target_encode('employee_residence', 'salary_in_usd') %>%
  target_encode('company_location', 'salary_in_usd') %>%
  target_encode('job_title', 'salary_in_usd') %>%
  target_encode('remote_ratio', 'salary_in_usd')

#matrix for validation data
test_data <- as.matrix(test_set_encoded %>% select(-salary_in_usd))

test_label <- test_set_encoded$salary_in_usd

test_preds <- predict(model, test_data)

mse3 <- mean((test_label - test_preds)^2)

rmse3 <- sqrt(mse3)

print(paste("Testing MSE: ", mse3))

print(paste("Testing RMSE: ", rmse3))

plot(test_label, test_preds, 
     xlab = "Actual Salary (USD)", 
     ylab = "Predicted Salary (USD)", 
     main = "Predicted vs Actual Salary",
     pch = 16, cex = 0.5, col = rgb(0,0,1,0.5))
abline(0, 1, col = "red", lwd = 2)
```

```{r}
lgb.importance(model, percentage = TRUE)
```

## 3. Results

Using the LightGBM model, I was able to predict salaries based on factors such as experience level, company size, employment type, employee residence, company location, job title, and remote ratio. The model achieved the following performance:

-   **Training RMSE**: \$3,730.67\
-   **Validation RMSE**: \$9,246.20\
-   **Testing RMSE**: \$8,274.13

These results suggest that while the model fits the training data well, it generalizes reasonably to unseen data, with an average prediction error of around \$8,274 on the test set. The predicted vs. actual salary plot showed a strong correlation, with points closely aligned along the diagonal line. This indicates that the model was generally able to capture the salary trends accurately across different employee profiles.

As expected, job title emerged as one of the most important predictors in the model. This makes sense, as roles like Research Engineer or Machine Learning Engineer often come with substantially higher average salaries. Other influential features included employee residence and experience level, reflecting the impact of geographic location and seniority on compensation. On the other hand, variables such as employment type and company size had relatively low importance in the model.
